---
execute:
  enabled: false
title: "Code, data & slides"
subtitle: "Everything you need to work through the examples."
toc-title: "ON THIS PAGE"
---

## Set up R

Use a current R release with RStudio. The examples are checked with R 4.5.3. Chapter 1 uses the `survival` package; Chapter 2 introduces the tidyverse and descriptive tables with `gtsummary`. Install packages once, then load them with `library()` in each new R session.

```{r}
#| eval: false
install.packages(c("survival", "tidyverse", "gtsummary", "gt"))
```

Later chapters use additional packages:

```{r}
#| eval: false
install.packages(c("broom", "ggsurvfit", "survminer", "tidycmprsk",
                   "tidymodels", "censored", "glmnet", "ranger", "broom.helpers"))
```

## Working data

The German breast cancer study is the running example. Download the data into a folder called `data` inside your RStudio project. The scripts use paths such as `data/gbc.txt` relative to that project folder.

| File | Contents |
|:---|:---|
| [gbc_mort.txt](data/gbc_mort.txt){download="gbc_mort.txt"} | Mortality follow-up, one record per patient |
| [gbc.txt](data/gbc.txt){download="gbc.txt"} | Relapse and mortality records; patients can contribute multiple rows |
| [gbc_ms.txt](data/gbc_ms.txt){download="gbc_ms.txt"} | Companion multistate data |
| [Data dictionary](data/Readme_gbc.txt) | Study provenance, variable descriptions, and coding |

Read the dictionary before choosing the outcome: mortality, relapse-free survival, and competing events answer different questions.

## Chapter materials

The scripts for Chapters 1 and 2 follow the book's worked examples and exercise solutions in order. The later scripts preserve the supplied chapter code and are provided for exploration as their narrative chapters develop.

| Chapter | Slides | R script |
|:---|:---|:---|
| 01 · Survival analysis in R | [PDF](slides/Module%201.pdf){target="_blank"} | [Download](code/chapter-1.R){download="chapter-1.R"} |
| 02 · Working with tidy data | [PDF](slides/Module%202.pdf){target="_blank"} | [Download](code/chapter-2.R){download="chapter-2.R"} |
| 03 · Nonparametric methods | [PDF](slides/Module%203.pdf){target="_blank"} | [Download](code/chapter-3.R){download="chapter-3.R"} |
| 04 · Regression models | [PDF](slides/Module%204.pdf){target="_blank"} | [Download](code/chapter-4.R){download="chapter-4.R"} |
| 05 · Machine learning | [PDF](slides/Module%205.pdf){target="_blank"} | [Download](code/chapter-5.R){download="chapter-5.R"} |

### Saved model fits

The machine-learning examples include saved tuning results. Put these in the same `data` folder if you want to inspect the supplied results without rerunning the tuning steps: [Cox tuning results](data/cox_res.rds) · [Random-forest tuning results](data/rf_res.rds).

## A note on the materials

The slides preserve the original short-course content and formatting. This book is being developed chapter by chapter; the supplied examples will be reviewed as each narrative chapter is expanded.
